---
title: "Consecutive digestion manuscript"
author: "Zandra Fagernäs
output: html_notebook
---

This notebook contains all analyses conducted for the manuscript "Digging deeper into ancient skeletal proteomes through consecutive digestion with multiple proteases". MaxQuant results and metadata can be found on ProteomeXchange (accession number will be added once published). The blanks were inspected during preliminary analysis, but are filtered out for statistics and figures. Probable contaminants were identified by going through the protein lists.

```{r }
library(MetBrewer) #Use Derain
library(janitor)
library(ggpubr)
library(venn)
library(lmerTest)
library(lme4)
library(MASS)
library(Hmisc)
library(vegan)
library(ggseqlogo)
library(tidyverse)

meta <- read_delim("<PATH_TO>/metadata.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() 
```

# Single-protease digestion

## Protein number

```{r}
# Load datasets and clean
protein_cow <- read_delim("<PATH_TO>/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+")) %>% 
  filter(str_detect(fasta_headers, "BOVIN")) %>%
  filter(!str_detect(fasta_headers, "LOC100848132|Keratin|A7YWU6_BOVIN|CTRB_BOVIN|CTRA_BOVIN|TRY1_BOVIN|A0A3Q1LWD8_BOVIN|protease")) %>%
  dplyr::select(fasta_headers, 16:20)
protein_horse <- read_delim("<PATH_TO>/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+"))  %>% 
  filter(str_detect(fasta_headers, "HORSE")) %>%
  filter(!str_detect(fasta_headers, "LOC100055297|Keratin|A0A3Q2IAH5_HORSE|F6T323_HORSE|LOC100146723|protease|A0A5F5PWX5_HORSE"))%>%
  dplyr::select(fasta_headers, 20:28)

# Long format, filter for identified proteins per sample, edit sample IDs
protein_cow <- protein_cow %>%
  gather(sample, peptides, 2:6) %>%
  filter(peptides > 0) %>%
  mutate(sample = gsub('razor_unique_peptides_', '', sample))
protein_horse <- protein_horse %>%
  gather(sample, peptides, 2:10) %>%
  filter(peptides > 0) %>%
  mutate(sample = gsub('razor_unique_peptides_', '', sample)) %>%
  filter(!sample == "blank")
  
# Combine datasets
protein <- full_join(protein_cow, protein_horse)

# Count proteins 
protein_summary <- protein %>%
  group_by(sample) %>%
  summarize(protein_count = n())

# Add metadata
protein_summary <- left_join(protein_summary, meta)

# Find Box-Cox transformation of the data
MASS::boxcox(lm(protein_count~bone, data=protein_summary))
MASS::boxcox(lm(protein_count~enzyme, data=protein_summary)) 

# Checking transformation
ggqqplot(sqrt(protein_summary$protein_count))
hist(sqrt(protein_summary$protein_count))

# Full model - start with full and decrease as needed
model1 <- lm(sqrt(protein_count) ~ enzyme, data = protein_summary)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE)

# Null model
model0 <- lm(sqrt(protein_count) ~ 1, data = protein_summary)

# Compare models
anova(model1, model0) 

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1)) 

# Plot
protein_count_fig <- ggplot(protein_summary, aes(x=sample, y=protein_count, fill=enzyme)) + 
  geom_bar(stat="identity") +
  facet_wrap(~bone, scales = "free_x") +
  scale_fill_manual(values=met.brewer("Derain", 5), name="Enzyme") +
  xlab("") +
  ylab("Protein count") +
  theme_bw() +
  theme(axis.text.x=element_blank())
```

## Non-collagenous proteins

```{r}
# Count NCPs
NCP_summary <- protein %>%
  filter(!str_detect(fasta_headers, "Collagen")) %>%
  group_by(sample) %>%
  summarize(ncp_count = n())

# Add protein count
NCP_summary <- left_join(NCP_summary, protein_summary)

# Calculate percent NCPs
NCP_summary$ncp_perc <- (NCP_summary$ncp_count / NCP_summary$protein_count) * 100

# Find Box-Cox transformation of the data
MASS::boxcox(lm(ncp_perc~enzyme, data=NCP_summary))
MASS::boxcox(lm(ncp_perc~bone+enzyme, data=NCP_summary))

# Checking transformation
ggqqplot(NCP_summary$ncp_perc)
hist(NCP_summary$ncp_perc)

# Full model - start with full and decrease as needed
model1 <- lm(ncp_perc ~ enzyme, data = NCP_summary)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE)

# Null model
model0 <- lm(ncp_perc ~ 1, data = NCP_summary)

# Compare models
anova(model1, model0)

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

# Plot
ncp_figure <- ggplot(NCP_summary, aes(x=sample, y=ncp_perc, fill=enzyme)) + 
  geom_bar(stat="identity") +
  facet_wrap(~bone, scales = "free_x") +
  scale_fill_manual(values=met.brewer("Derain", 5)) +
  xlab("") +
  ylab("NCP %") +
  theme_bw() +
  theme(axis.text.x=element_blank())
```

## Peptide number

```{r}
# Count peptides
peptide_summary <- protein %>%
  group_by(sample) %>%
  summarize(peptide_count = sum(peptides))

# Add metadata
peptide_summary <- left_join(peptide_summary, meta)

# Find Box-Cox transformation of the data
MASS::boxcox(lm(peptide_count~bone, data=peptide_summary))
MASS::boxcox(lm(peptide_count~enzyme, data=peptide_summary))

# Checking transformation
ggqqplot(sqrt(peptide_summary$peptide_count)) # Good
hist(sqrt(peptide_summary$peptide_count)) # Not great, but very few observations

# Full model - start with full and decrease as needed
model1 <- lm(sqrt(peptide_count) ~ enzyme, data = peptide_summary)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE)

# Null model
model0 <- lm(sqrt(peptide_count) ~ 1, data = peptide_summary)

# Compare models
anova(model1, model0)

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1)) 

# Plot
peptide_count_fig <- ggplot(peptide_summary, aes(x=sample, y=peptide_count, fill=enzyme)) + 
  geom_bar(stat="identity") +
  facet_wrap(~bone, scales = "free_x") +
  scale_fill_manual(values=met.brewer("Derain", 5), name="Enzyme") +
  xlab("") +
  ylab("Peptide count") +
  theme_bw() +
  theme(axis.text.x=element_blank())

# Combine figures
proteome_fig <- ggarrange(protein_count_fig, peptide_count_fig, length,
          common.legend=TRUE, nrow=1, labels=c("a)", "b)", "c)"), legend="bottom")

# Export pdf
ggsave("<PATH_TO>/enzexp1_proteome.pdf", plot=proteome_fig, width=12, height=4)
```

## PCA

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load datasets and clean
gene_cow <- read_delim("<PATH_TO>/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+")) %>% 
  filter(str_detect(fasta_headers, "BOVIN")) %>%
  filter(!str_detect(fasta_headers, "LOC100848132|Keratin|A7YWU6_BOVIN|CTRB_BOVIN|CTRA_BOVIN|TRY1_BOVIN|A0A3Q1LWD8_BOVIN|protease")) %>%
  dplyr::select(fasta_headers, 47:51)
gene_horse <- read_delim("<PATH_TO>/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+"))  %>% 
  filter(str_detect(fasta_headers, "HORSE")) %>%
filter(!str_detect(fasta_headers, "LOC100055297|Keratin|A0A3Q2IAH5_HORSE|F6T323_HORSE|LOC100146723|protease|A0A5F5PWX5_HORSE")) %>%
  dplyr::select(fasta_headers, 67:75)

# Extract gene names
gene_cow <- gene_cow %>%
  separate(fasta_headers, c("remove", "gene"), "GN=") %>%
  separate(gene, c("gene", "remove2"), " PE=") %>%
  dplyr::select(-remove, -remove2) %>%
  filter(!is.na(gene))
gene_horse <- gene_horse %>%
  separate(fasta_headers, c("remove", "gene"), "GN=") %>%
  separate(gene, c("gene", "remove2"), " PE=") %>%
  dplyr::select(-remove, -remove2) %>%
  filter(!is.na(gene))

# Combine datasets
genes_all <- full_join(gene_cow, gene_horse) %>%
  mutate(across(where(is.numeric), ~ replace_na(.x, 0)))

# Filter for presence in two samples 
genes_all$presence <- rowSums(genes_all[2:15] != 0)
genes_all <- genes_all %>%
  filter(presence > 1) %>%
  dplyr::select(-presence)

# Transpose
genes_all <- genes_all %>%
  gather(sample, intensity, 2:15) %>%
  spread(gene, intensity) %>%
  replace(is.na(.), 0)

# Fix sample names
genes_all <- genes_all %>%
  mutate(sample = gsub('intensity_', '', sample))

# Pseudocount zero replacement
genes_all[,2:ncol(genes_all)] <- genes_all[,2:ncol(genes_all)]+1

# Log-transformation
genes_all[,2:ncol(genes_all)] <- log2(genes_all[,2:ncol(genes_all)])

# Create PCA object
pca_all <- prcomp(genes_all[, 2:255], scale=FALSE)

# Extract values
pca_out <- as_tibble(pca_all$x)

# Add metadata
pca_out$sample <- genes_all$sample
pca_out <- left_join(pca_out, meta)

# Calculate percent explained per PC
percentage <- round(pca_all$sdev / sum(pca_all$sdev) * 100, 2)
percentage <- paste(colnames(pca_out), "(", paste(
  as.character(percentage), "%", ")", sep="") )

# Get variables with biggest impact for PC1 and PC2
pca_all$rotation[,1] %>% sort() %>% head(1)
pca_all$rotation[,1] %>% sort(decreasing = TRUE) %>% head(1)
pca_all$rotation[,2] %>% sort() %>% head(1)
pca_all$rotation[,2] %>% sort(decreasing = TRUE) %>% head(1)

# Get rownames for each protein
which(rownames(pca_all$rotation) %in% 
        c("COL22A1", "F2", "COL12A1", "ZFHX4"))

# Create matrix of x (PC1) coordinates and y (PC2) 
l.x <- cbind(pca_all$rotation[,1][c(45,52,94,246)]) * 500
l.y <- cbind(pca_all$rotation[,2][c(45,52,94,246)]) * 500

# Plot
pca <- ggplot(data = pca_out, aes(x = PC1, y = PC2, fill=enzyme, shape=bone)) +
  geom_point(size=6, stroke=1) +
  xlab(percentage[1]) + 
  ylab(percentage[2]) +
  scale_fill_manual(values=met.brewer(n=5, name="Derain")) +
  scale_shape_manual(values=c(21,22,23)) +
  guides(fill=guide_legend(override.aes=list(shape=21, size=5))) +
  theme_minimal(base_size = 10) +
  annotate("segment", x=0, xend=l.x, y=0, yend=l.y, colour="black", 
           size=1, arrow=arrow()) + 
  annotate("text", x=l.x, y=l.y, label=rownames(l.x), color="black")

# Export plot
ggsave("<PATH_TO>/enzexp1_pca.pdf", plot=pca, width=8, height=6)

# Basic permanova with euclidean distance as metric
pca_meta <- left_join(genes_all, meta) %>%
  dplyr::select(sample, enzyme, bone, age, reference)
permanova <- adonis(genes_all[, 2:255] ~ enzyme + age, 
                    data=pca_meta,
                    permutations=999,
                    method="euclidean")
permanova

# Checking assumptions
dist <- vegdist(genes_all[, 2:255])
anova(betadisper(dist, pca_meta$bone))
```

## Protein overlap

```{r}
# Add metadata
protein_venn <- left_join(protein, meta)

# Filter for samples separately, select columns
sct_venn <- protein_venn %>%
  filter(bone=="SCT9") %>%
  filter(peptides > 0) %>%
  dplyr::select(fasta_headers, enzyme)
lf_venn <- protein_venn %>%
  filter(bone=="LF59") %>%
  filter(peptides > 0) %>%
  dplyr::select(fasta_headers, enzyme)
grl_venn <- protein_venn %>%
  filter(bone=="GRL12497") %>%
  filter(peptides > 0) %>%
  dplyr::select(fasta_headers, enzyme)

# Make list
sct_venn <- list(
  trypsin = sct_venn %>% filter(enzyme == "Trypsin") %>% pull(fasta_headers),
  chymo = sct_venn %>% filter(enzyme == "Chymotrypsin") %>%pull(fasta_headers),
  glu = sct_venn %>% filter(enzyme == "Glu-C") %>%pull(fasta_headers),
  lys = sct_venn %>% filter(enzyme == "LysargiNase") %>%pull(fasta_headers),
  pro = sct_venn %>% filter(enzyme == "ProAlanase") %>%pull(fasta_headers))
lf_venn <- list(
  trypsin = lf_venn %>% filter(enzyme == "Trypsin") %>% pull(fasta_headers),
  chymo = lf_venn %>% filter(enzyme == "Chymotrypsin") %>%pull(fasta_headers),
  glu = lf_venn %>% filter(enzyme == "Glu-C") %>%pull(fasta_headers),
  lys = lf_venn %>% filter(enzyme == "LysargiNase") %>%pull(fasta_headers),
  pro = lf_venn %>% filter(enzyme == "ProAlanase") %>%pull(fasta_headers))
grl_venn <- list(
  trypsin = grl_venn %>% filter(enzyme == "Trypsin") %>% pull(fasta_headers),
  chymo = grl_venn %>% filter(enzyme == "Chymotrypsin") %>%pull(fasta_headers),
  glu = grl_venn %>% filter(enzyme == "Glu-C") %>%pull(fasta_headers),
  pro = grl_venn %>% filter(enzyme == "ProAlanase") %>%pull(fasta_headers))

# Plot
sct_venn_fig <- venn(sct_venn, ilables = "counts", zcolor=met.brewer("Derain", 5))
lf_venn_fig <- venn(lf_venn, ilables = "counts", zcolor=met.brewer("Derain", 5))
grl_venn_fig <- venn(grl_venn, ilables = "counts", zcolor=met.brewer("Derain", 5))
```

## Protein coverage

COL1A1 - P02453 and A0A5F5Q281.
AHSG/FETUA - P12763 and A0A5F5Q0V8.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load peptide files
pep_cow <- read_delim("<PATH_TO>/peptides.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(str_detect(proteins, "P12763")) %>%
  dplyr::select(62:66, start_position, end_position)
pep_horse <- read_delim("<PATH_TO>/peptides.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(str_detect(proteins, "A0A5F5Q0V8")) %>%
  dplyr::select(70:78, start_position, end_position)

# Fix sample names
colnames(pep_cow)[1:5] <- c("sct_chy", "sct_glu", "sct_lys", "sct_pro", "sct_try")
colnames(pep_horse)[1:9] <- c("grl_chy", "grl_glu", "grl_pro", "grl_try","lf_chy", "lf_glu", "lf_lys", "lf_pro", "lf_try")

# Combine datasets
pep_all <- full_join(pep_cow, pep_horse) %>%
  dplyr::select(start_position, end_position, 1:5, 8:16)

# Expand numbers between start and end to get a dataframe of all covered positions
pep_all <- pep_all %>%
  mutate(covered = map2(start_position, end_position, seq, by=1))  
pos_list <- pep_all %>% dplyr::select(covered) %>%
  mutate(covered = map(covered, ~as_data_frame(t(.x)))) %>% unnest()
pep_all <- cbind(pep_all, pos_list)

# Gather to long form
pep_all <- pep_all %>%
  gather("remove", "position", 18:37) %>% 
  dplyr::select(-start_position, -end_position, -covered, -remove) %>%
  filter(position > 0) %>%
  gather("sample", "intensity", 1:14) %>%
  filter(intensity > 0)

# Sum intensity per position
pep_summary <- pep_all %>%
  group_by(position, sample) %>%
  summarize(total_intensity = sum(intensity))

# Add metadata
pep_summary <- left_join(pep_summary, meta) %>%
  dplyr::select(position, sample, total_intensity, enzyme, bone)

# log2 intensity
pep_summary$intensity_log2 <- log2(pep_summary$total_intensity)

# Separate into species (cut collagen only)
pep_summary_cow <- pep_summary %>%
  filter(bone == "SCT9") 
  filter(position < 1217)
pep_summary_horse <- pep_summary %>%
  filter(bone %in% c("GRL12497", "LF59"))
  filter(position < 1184)

# Reorder for plotting
pep_summary_cow$sample <- factor(pep_summary_cow$sample, 
                                 levels=c("sct_try", "sct_chy", "sct_glu", "sct_lys", "sct_pro"))
pep_summary_horse$sample <- factor(pep_summary_horse$sample, levels=c("grl_try", "lf_try",
                                                          "grl_chy", "lf_chy",
                                                          "grl_glu", "lf_glu",
                                                          "lf_lys",
                                                          "grl_pro", "lf_pro"))


# Plot
fetua_cow <- ggplot(pep_summary_cow, aes(x=position, y=sample)) +
  geom_raster(aes(fill=intensity_log2)) +
  scale_fill_gradient(low="#e9eaf2ff", high="#6e9867ff", name="Intensity (log2)") +
  xlab("") + ylab("") +
  theme_bw()

# Combine plots
col1a1 <- ggarrange(col1a1_cow, col1a1_horse, common.legend=TRUE, ncol=1, heights=c(3,4))
fetua <- ggarrange(fetua_cow, fetua_horse, common.legend=TRUE, ncol=1)
coverage <- ggarrange(col1a1, fetua, labels=c("a)", "b)"), common.legend=TRUE, widths = c(2,1))

# Export plot
ggsave("<PATH_TO>/enzexp1_coverage.pdf", coverage, width=10, height=4)
```

## Peptide length

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load data
evidence_cow<- read_delim("<PATH_TO>/evidence.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "F1N152|A0A452DI93|P00766|A0A3Q1LWD8|Q29S21|Q08D91|A5D7M6|Q32LI2|P0C1U8|ENSBTAP00000018229|P04264|P00761|P07477|P0C1U8")) %>%
  dplyr::select(proteins, length, experiment, intensity) 
evidence_horses <- read_delim("<PATH_TO>/evidence.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "F6T323|F6S8W6|A0A3Q2LCW0|F7C210|A0A5F5PWX5|F6TW38|F7C2N1|P0C1U8|P00761|P00766|P04264|F7BC37|P13645|P35527|P35908|P07477")) %>%
  dplyr::select(proteins, length, experiment, intensity) 

# Combine files
evidence <- full_join(evidence_cow, evidence_horses) %>%
  filter(intensity > 0) %>%
  filter(!experiment == "blank")

# Add metadata
evidence <- left_join(evidence, meta, by=c("experiment"="sample"))

# Summarize
length_summary <- evidence %>%
  group_by(experiment, enzyme, bone, age, reference) %>%
  dplyr::summarise(mean_length = weighted.mean(x=length, w=intensity, na.rm=TRUE),
                   weighted_var= wtd.var(x=length, w=intensity, na.rm=TRUE))
length_summary$weighted_sd <- sqrt(length_summary$weighted_var)

# Find Box-Cox transformation of the data
MASS::boxcox(lm(mean_length~bone, data=length_summary))
MASS::boxcox(lm(mean_length~enzyme, data=length_summary))

# Checking transformation
ggqqplot(1/(length_summary$mean_length)) 
hist(1/(length_summary$mean_length)) 

# Full model - start with full and decrease as needed
model1 <- lmer(1/(mean_length) ~ enzyme + (1|age), data = length_summary)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) 

# Null model
model0 <- lmer(1/(mean_length) ~ 1 + (1|age), data = length_summary)

# Compare models
anova(model1, model0) 

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

# Plot
length <- ggplot(length_summary, aes(x=experiment, y=mean_length, fill=enzyme)) + 
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin=mean_length-weighted_sd, ymax=mean_length+weighted_sd), width=.2,
                 position=position_dodge(.9)) +
  facet_wrap(~bone, scales = "free_x") +
  scale_fill_manual(values=met.brewer("Derain", 5), name="Enzyme") +
  xlab("") +
  ylab("Mean length weighted by intensity") +
  theme_bw() +
  theme(axis.text.x=element_blank())
```

## Hydroxyprolines

```{r}
# Load and clean evidence files
evidence_cow <- read_delim("<PATH_TO>/evidence.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>% 
  filter(!reverse %in% c("+")) %>% 
  filter(!str_detect(proteins, "F1N152|A0A452DI93|P00766|A0A3Q1LWD8|Q29S21|Q08D91|A5D7M6|Q32LI2|P0C1U8|ENSBTAP00000018229|P04264|P00761|P07477|P0C1U8")) %>%
  select(sequence, experiment, intensity, modified_sequence, leading_proteins,
         deamidation_nq, oxidation_m, hydroxyproline, gln_pyro_glu, glu_pyro_glu)
evidence_horse <- read_delim("<PATH_TO>/evidence.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>% 
  filter(!reverse %in% c("+")) %>% 
  filter(!str_detect(proteins, "F6T323|F6S8W6|A0A3Q2LCW0|F7C210|A0A5F5PWX5|F6TW38|F7C2N1|P0C1U8|P00761|P00766|P04264|F7BC37|P13645|P35527|P35908|P07477")) %>%
  select(sequence, experiment, intensity, modified_sequence, leading_proteins,
         deamidation_nq, oxidation_m, hydroxyproline, gln_pyro_glu, glu_pyro_glu)

# Reduce to only hydroxyprolines
hydro_cow <- evidence_cow %>%
  dplyr::select(sequence, hydroxyproline, experiment)
hydro_horses <- evidence_horse %>%
  dplyr::select(sequence, hydroxyproline, experiment)

# Combine
hydro <- full_join(hydro_cow, hydro_horses)

# Count number of P in peptide
hydro$count_p <- str_count(hydro$sequence, "P")

# Remove peptides with no P
hydro <- hydro %>% filter(count_p > 0)

# Fraction hydroxyproline
hydro$fraction_hydro <- hydro$hydroxyproline / hydro$count_p

# Mean fraction hydroxyproline by sample
hydro_summary <- hydro %>%
  group_by(experiment) %>%
  dplyr::summarize(mean_hydro = mean(fraction_hydro))

# Add metadata
hydro_summary <- left_join(hydro_summary, meta, by=c("experiment"="sample")) %>%
  filter(!experiment == "blank")

# Find Box-Cox transformation of the data
MASS::boxcox(lm(mean_hydro~age, data=hydro_summary))
MASS::boxcox(lm(mean_hydro~enzyme, data=hydro_summary)) 

# Checking transformation
ggqqplot(1/(hydro_summary$mean_hydro)) 
hist(1/(hydro_summary$mean_hydro)) 

# Full model - start with full and decrease as needed 
model1 <- lm(1/mean_hydro ~ enzyme, data = hydro_summary)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) 

# Null model
model0 <- lm(1/mean_hydro ~ 1, data = hydro_summary)

# Compare models
anova(model1, model0)

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

# Plot
ggplot(hydro_summary, aes(x=bone, y=mean_hydro, fill=enzyme)) + 
  geom_bar(stat="identity", position="dodge") +
  scale_fill_manual(values=met.brewer("Derain", 5), name="Enzyme") +
  xlab("") +
  theme_bw()  +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Deamidation

This analysis is based on output from the deamidate.py script published by Mackie et al. 2018. Input was cleaned evidence.txt files.

```{r}
# Load results
deam <- read.csv("<PATH_TO>/enzexp1_deamidation.txt", sep="") %>%
  clean_names()

# Extract stagetip number
deam <- deam %>%
  mutate(raw_file = gsub('_Bone', '', raw_file)) %>%
  separate(raw_file, c("remove", "stagetip"), sep="DDA_") %>%
  select(-remove)

# Add metadata
deam <- left_join(deam, meta)

# Separate into N and Q for stats
deam_n <- deam %>% filter(n_q == "N")
deam_q <- deam %>% filter(n_q == "Q")

### Stats N

# Find Box-Cox transformation of the data
MASS::boxcox(lm(mean~bone, data=deam_n))
MASS::boxcox(lm(mean~enzyme, data=deam_n)) 

# Checking transformation
ggqqplot(log(deam_n$mean)) 
hist(log(deam_n$mean)) 

# Full model - start with full and decrese as needed
model1 <- lmer(log(mean) ~ enzyme + (1|age), data = deam_n)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) 

# Null model
model0 <- lmer(log(mean) ~ 1 + (1|age), data = deam_n)

# Compare models
anova(model1, model0)

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

### Stats Q

# Find Box-Cox transformation of the data
MASS::boxcox(lm(mean~bone, data=deam_q))
MASS::boxcox(lm(mean~enzyme, data=deam_q)) 

# Checking transformation
ggqqplot(deam_q$mean^2) 
hist(deam_q$mean^2) 

# Full model - start with full and decrese as needed
model1 <- lmer(mean^2 ~ enzyme + (1|age), data = deam_q)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) 

# Null model
model0 <- lmer(mean^2 ~ 1 + (1|age), data = deam_q)

# Compare models
anova(model1, model0)

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

# Plot
deamidation_fig <- ggplot(deam, aes(x=n_q, y=mean, fill=enzyme)) + 
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin=mean-std, ymax=mean+std), width=.2,
                 position=position_dodge(.9)) +
  scale_fill_manual(values=met.brewer("Derain", 5), name="Enzyme") +
  facet_grid(~bone) +
  xlab("") + 
  ylab("Deamidation") +
  theme_bw() 

# Combine figures
damage <- ggarrange(length, deamidation_fig, common.legend=TRUE, 
                    labels=c("a)", "b)"), widths=c(3,4))

# Export figure
ggsave("<PATH_TO>/enzexp1_damage.pdf", damage, width=9, height=4)
```

## Cleavage sites

```{r}
# Import datasets 
peptide_cow <- read_delim("<PATH_TO>/peptides.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "F1N152|A0A452DI93|P00766|A0A3Q1LWD8|Q29S21|Q08D91|A5D7M6|Q32LI2|P0C1U8|ENSBTAP00000018229|P04264|P00761|P07477|P0C1U8")) %>%
  select("n_term_cleavage_window", 62:66)
peptide_horse <- read_delim("<PATH_TO>/peptides.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "F6T323|F6S8W6|A0A3Q2LCW0|F7C210|A0A5F5PWX5|F6TW38|F7C2N1|P0C1U8|P00761|P00766|P04264|F7BC37|P13645|P35527|P35908|P07477")) %>%
  select("n_term_cleavage_window", 70:78)

# Remove peptides with missing aa's
peptide_cow <- peptide_cow %>%
  filter(!grepl('_', n_term_cleavage_window))
peptide_horse <- peptide_horse %>%
  filter(!grepl('_', n_term_cleavage_window))

# Filter for  presence
peptide_cow <- peptide_cow %>%
  pivot_longer(names_to = "enzyme", values_to = "intensity", 2:6) %>%
  filter(intensity >0) %>%
  separate(enzyme, c("remove", "sample"), "sity_") %>%
  select(-remove)
peptide_horse <- peptide_horse %>%
  pivot_longer(names_to = "enzyme", values_to = "intensity", 2:10) %>%
  filter(intensity >0) %>%
  separate(enzyme, c("remove", "sample"), "sity_") %>%
  select(-remove)

# Cut first and last 10 aa's
peptide_cow <- peptide_cow %>%
  mutate(n_term_cleavage_window = str_sub(n_term_cleavage_window, 11, -1)) %>%
  mutate(n_term_cleavage_window = str_sub(n_term_cleavage_window, end=-11))
peptide_horse <- peptide_horse %>%
  mutate(n_term_cleavage_window = str_sub(n_term_cleavage_window, 11, -1)) %>%
  mutate(n_term_cleavage_window = str_sub(n_term_cleavage_window, end=-11))

# Add metadata
peptide_cow <- left_join(peptide_cow, meta)
peptide_horse <- left_join(peptide_horse, meta)

# Summarize
peptide_cow_summary <- peptide_cow %>%
  select(n_term_cleavage_window, enzyme, bone) %>%
  distinct()
peptide_horse_summary <- peptide_horse %>%
  select(n_term_cleavage_window, enzyme, bone) %>%
  distinct()

# Make into list - there's gotta be a more smooth way to do this but I'm tired
peptide_cow_trypsin <- peptide_cow_summary %>%
  filter(enzyme == "Trypsin")
peptide_cow_chy <- peptide_cow_summary %>%
  filter(enzyme == "Chymotrypsin")
peptide_cow_glu <- peptide_cow_summary %>%
  filter(enzyme == "Glu-C")
peptide_cow_pro <- peptide_cow_summary %>%
  filter(enzyme == "ProAlanase")
peptide_cow_lys <- peptide_cow_summary %>%
  filter(enzyme == "LysargiNase")
peptide_cow_list <- list(peptide_cow_trypsin$n_term_cleavage_window,
                       peptide_cow_chy$n_term_cleavage_window,
                       peptide_cow_glu$n_term_cleavage_window,
                       peptide_cow_pro$n_term_cleavage_window,
                       peptide_cow_lys$n_term_cleavage_window)
names(peptide_cow_list) <- c("Trypsin", "Chymotrypsin", "Glu-C", "ProAlanase", "LysargiNase") 

peptide_horse_trypsin <- peptide_horse_summary %>%
  filter(enzyme == "Trypsin")
peptide_horse_chy <- peptide_horse_summary %>%
  filter(enzyme == "Chymotrypsin")
peptide_horse_glu <- peptide_horse_summary %>%
  filter(enzyme == "Glu-C")
peptide_horse_pro <- peptide_horse_summary %>%
  filter(enzyme == "ProAlanase")
peptide_horse_lys <- peptide_horse_summary %>%
  filter(enzyme == "LysargiNase")
peptide_horse_list <- list(peptide_horse_trypsin$n_term_cleavage_window,
                       peptide_horse_chy$n_term_cleavage_window,
                       peptide_horse_glu$n_term_cleavage_window,
                       peptide_horse_pro$n_term_cleavage_window,
                       peptide_horse_lys$n_term_cleavage_window)
names(peptide_horse_list) <- c("Trypsin", "Chymotrypsin", "Glu-C", "ProAlanase", "LysargiNase") 

# Plot
seq_cow <- ggseqlogo(peptide_cow_list, ncol=2)
seq_horse <- ggseqlogo(peptide_horse_list, ncol=2)

seq_logos <- ggarrange(seq_cow, seq_horse, nrow=1, common.legend=TRUE, legend="bottom",
          labels=c("a)", "b)"))
```


# Consecutive digestion

## Protein number

```{r}
# Load datasets and clean
protein_horse <- read_delim("<PATH_TO>/combined/txt/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+"))  %>% 
  filter(str_detect(fasta_headers, "HORSE")) %>%
  filter(!str_detect(fasta_headers, "LOC100055297|Keratin|Trypsin|trypsinogen|protease")) %>%
  dplyr::select(fasta_headers, 33:54)
protein_cow <- read_delim("<PATH_TO>/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+"))  %>% 
  filter(str_detect(fasta_headers, "BOVIN")) %>%
  filter(!str_detect(fasta_headers, "LOC100848132|Keratin|A7YWU6_BOVIN|CTRB_BOVIN|CTRA_BOVIN|TRY1_BOVIN|Trypsin|trypsinogen|protease")) %>%
  dplyr::select(fasta_headers, 21:30)

# Long format, filter for identified proteins per sample, edit sample IDs
protein_horse <- protein_horse %>%
  gather(sample, peptides, 2:23) %>%
  filter(peptides > 0) %>%
  mutate(sample = gsub('peptides_', '', sample))
protein_cow <- protein_cow %>%
  gather(sample, peptides, 2:11) %>%
  filter(peptides > 0) %>%
  mutate(sample = gsub('peptides_', '', sample))

# Combine datasets
protein <- full_join(protein_horse, protein_cow)

# Count proteins 
protein_summary <- protein %>%
  group_by(sample) %>%
  dplyr::summarize(protein_count = n()) %>%
  mutate(sample = gsub('razor_unique_', '', sample))

# Add metadata
protein_summary <- left_join(protein_summary, meta)

# Find Box-Cox transformation of the data
MASS::boxcox(lm(protein_count~bone, data=protein_summary))
MASS::boxcox(lm(protein_count~enzyme, data=protein_summary))

# Checking transformation
ggqqplot(log(protein_summary$protein_count))
hist(log(protein_summary$protein_count))

# Full model - start with full and decrease as needed
model1 <- lmer(log(protein_count) ~ enzyme + (1|age), data = protein_summary)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) 

# Null model
model0 <- lmer(log(protein_count) ~ 1 + (1|age), data = protein_summary)

# Compare models
anova(model1, model0) 

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

# Plot
protein_count <- ggplot(protein_summary, aes(x=enzyme, y=protein_count, fill=enzyme)) + 
  geom_boxplot() +
  stat_summary(fun.y=median, geom="point", shape=21, size=3.5) +
  facet_wrap(~bone) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_fill_manual(values=met.brewer("Derain", 5), name="Enzyme") +
  xlab("") +
  ylab("Protein count") +
  theme_bw() +
  theme(axis.text.x=element_blank())
```

## Non-collagenous proteins

```{r}
# Count NCPs
NCP_summary <- protein %>%
  filter(!str_detect(fasta_headers, "Collagen")) %>%
  group_by(sample) %>%
  dplyr::summarize(ncp_count = n()) %>%
  mutate(sample = gsub('razor_unique_', '', sample))

# Add protein count
NCP_summary <- left_join(NCP_summary, protein_summary)

# Calculate percent NCPs
NCP_summary$ncp_perc <- (NCP_summary$ncp_count / NCP_summary$protein_count) * 100

# Find Box-Cox transformation of the data
MASS::boxcox(lm(ncp_perc~enzyme, data=NCP_summary))
MASS::boxcox(lm(ncp_perc~bone+enzyme, data=NCP_summary))

# Checking transformation
ggqqplot(log(NCP_summary$ncp_perc)) 
hist(log(NCP_summary$ncp_perc)) 

# Full model - start with full and decrease as needed
model1 <- lmer(log(ncp_perc) ~ enzyme + (1|bone), data = NCP_summary)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) # None needed

# Null model
model0 <- lmer(log(ncp_perc) ~ 1 + (1|bone), data = NCP_summary)

# Compare models
anova(model1, model0) 

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

# Plot
ggplot(NCP_summary, aes(x=enzyme, y=ncp_perc, fill=enzyme)) + 
  geom_boxplot() +
  facet_wrap(~bone, scales = "free_x") +
  scale_fill_manual(values=met.brewer("Derain", 5)) +
  xlab("") +
  ylab("NCP %") +
  theme_bw() +
  theme(axis.text.x=element_blank())
```

## Peptide number

```{r}
# Count peptides
peptide_summary <- protein %>%
  group_by(sample) %>%
  dplyr::summarize(peptide_count = sum(peptides)) %>%
  mutate(sample = gsub('razor_unique_', '', sample))

# Add metadata
peptide_summary <- left_join(peptide_summary, meta)

# Find Box-Cox transformation of the data
MASS::boxcox(lm(peptide_count~bone, data=peptide_summary))
MASS::boxcox(lm(peptide_count~enzyme, data=peptide_summary)) 

# Checking transformation
ggqqplot(log(peptide_summary$peptide_count))
hist(log(peptide_summary$peptide_count))

# Full model - start with full and decrease as needed
model1 <- lmer(log(peptide_count) ~ enzyme + (1|age), data = peptide_summary)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) 

# Null model
model0 <- lmer(log(peptide_count) ~ 1 + (1|age), data = peptide_summary)

# Compare models
anova(model1, model0) 

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

# Plot
peptide_count <- ggplot(peptide_summary, aes(x=enzyme, y=peptide_count, fill=enzyme)) + 
  geom_boxplot() +
  stat_summary(fun.y=median, geom="point", shape=21, size=3.5) +
  facet_wrap(~bone) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_fill_manual(values=met.brewer("Derain", 5), name="Enzyme") +
  xlab("") +
  ylab("Peptide count") +
  theme_bw() +
  theme(axis.text.x=element_blank())

# Combine plots
count_fig <- ggarrange(protein_count, peptide_count,
          common.legend=TRUE, nrow=1, labels=c("a)", "b)"), legend="bottom")

# Export figure
ggsave("<PATH_TO>/enzexp2_protein_peptide_count.pdf", count_fig, width=8, height=4)
```

## PCA

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load datasets and clean
gene_horse <- read_delim("<PATH_TO>/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+"))  %>% 
  filter(str_detect(fasta_headers, "HORSE")) %>%
  filter(!str_detect(fasta_headers, "LOC100055297|Keratin|Trypsin|trypsinogen|protease")) %>%
  dplyr::select(fasta_headers, 132:153)
gene_cow <- read_delim("<PATH_TO>/proteinGroups.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names()  %>% 
  filter(!reverse %in% c("+"))  %>% 
  filter(str_detect(fasta_headers, "BOVIN")) %>%
  filter(!str_detect(fasta_headers, "LOC100848132|Keratin|A7YWU6_BOVIN|CTRB_BOVIN|CTRA_BOVIN|TRY1_BOVIN|Trypsin|trypsinogen|protease")) %>%
  dplyr::select(fasta_headers, 72:81)

# Extract gene names
gene_horse <- gene_horse %>%
  separate(fasta_headers, c("remove", "gene"), "GN=") %>%
  separate(gene, c("gene", "remove2"), " PE=") %>%
  dplyr::select(-remove, -remove2) %>%
  filter(!is.na(gene))
gene_cow <- gene_cow %>%
  separate(fasta_headers, c("remove", "gene"), "GN=") %>%
  separate(gene, c("gene", "remove2"), " PE=") %>%
  dplyr::select(-remove, -remove2) %>%
  filter(!is.na(gene))

# Combine datasets
genes_all <- full_join(gene_cow, gene_horse) %>%
  mutate(across(where(is.numeric), ~ replace_na(.x, 0)))

# Filter for presence in two samples 
genes_all$presence <- rowSums(genes_all[2:33] != 0)
genes_all <- genes_all %>%
  filter(presence > 1) %>%
  dplyr::select(-presence)

# Transpose
genes_all <- genes_all %>%
  gather(sample, intensity, 2:33) %>%
  group_by(sample, gene) %>%
  dplyr::summarize(intensity = sum(intensity)) %>%
  ungroup() %>%
  spread(gene, intensity) %>%
  replace(is.na(.), 0)

# Fix sample names
genes_all <- genes_all %>%
  mutate(sample = gsub('intensity_', '', sample))

# Pseudocount zero replacement
genes_all[,2:ncol(genes_all)] <- genes_all[,2:ncol(genes_all)]+1

# Log-transformation
genes_all[,2:ncol(genes_all)] <- log2(genes_all[,2:ncol(genes_all)])

# Create PCA object
pca_all <- prcomp(genes_all[, 2:301], scale=FALSE)

# Extract values
pca_out <- as_tibble(pca_all$x)

# Add metadata
pca_out$sample <- genes_all$sample
pca_out <- left_join(pca_out, meta)

# Calculate percent explained per PC
percentage <- round(pca_all$sdev / sum(pca_all$sdev) * 100, 2)
percentage <- paste(colnames(pca_out), "(", paste(
  as.character(percentage), "%", ")", sep="") )

# Get variables with biggest impact for PC1 and PC2
pca_all$rotation[,1] %>% sort() %>% head(1)
pca_all$rotation[,1] %>% sort(decreasing = TRUE) %>% head(1)
pca_all$rotation[,2] %>% sort() %>% head(1)
pca_all$rotation[,2] %>% sort(decreasing = TRUE) %>% head(1)

# Get rownames for each protein
which(rownames(pca_all$rotation) %in% 
        c("COLEC12", "MGP", "CHERP", "COL19A1"))

# Create matrix of x (PC1) coordinates and y (PC2) (duplicates removed)
l.x <- cbind(pca_all$rotation[,1][c(47,60,81,185)]) * 300
l.y <- cbind(pca_all$rotation[,2][c(47,60,81,185)]) * 300

# Plot
pca <- ggplot(data = pca_out, aes(x = PC1, y = PC2, fill=enzyme, shape=bone)) +
  geom_point(size=6, stroke=1) +
  xlab(percentage[1]) + 
  ylab(percentage[2]) +
  scale_fill_manual(values=met.brewer(n=5, name="Derain")) +
  scale_shape_manual(values=c(21,22,23)) +
  guides(fill=guide_legend(override.aes=list(shape=21, size=5))) +
  theme_minimal(base_size = 10) +
  annotate("segment", x=0, xend=l.x, y=0, yend=l.y, colour="black", 
           size=1, arrow=arrow()) + 
  annotate("text", x=l.x, y=l.y, label=rownames(l.x), color="black")

# Export plot
ggsave("I:/SCIENCE-SNM-PP_priv/Users/Zandra/Manuscripts/enzexp/enzexp2_pca_20231114.pdf", pca, width=8, height=6)

# Basic permanova with euclidean distance as metric
pca_meta <- left_join(genes_all, meta) %>%
  dplyr::select(sample, enzyme, bone, age)
permanova <- adonis(genes_all[, 2:301] ~ enzyme + age + bone, 
                    data=pca_meta,
                    permutations=999,
                    method="euclidean")
permanova

# Checking assumptions
dist <- vegdist(genes_all[, 2:301])
anova(betadisper(dist, pca_meta$age))
```

## Protein overlap

```{r}
# Add metadata
protein_venn <- protein %>%
  mutate(sample = gsub('razor_unique_', '', sample))
protein_venn <- left_join(protein_venn, meta) 

# Filter for samples separately, select columns
sct_venn <- protein_venn %>%
  filter(bone=="SCT9") %>%
  filter(peptides > 0) %>%
  dplyr::select(fasta_headers, enzyme)
f35_venn <- protein_venn %>%
  filter(bone=="F35") %>%
  filter(peptides > 0) %>%
  dplyr::select(fasta_headers, enzyme)
grl_venn <- protein_venn %>%
  filter(bone=="GRL12497") %>%
  filter(peptides > 0) %>%
  dplyr::select(fasta_headers, enzyme)

# Make list
sct_venn <- list(
  chy_try = sct_venn %>% filter(enzyme == "Chymotrypsin+trypsin") %>% pull(fasta_headers),
  glu_try = sct_venn %>% filter(enzyme == "Glu-C+trypsin") %>%pull(fasta_headers),
  try = sct_venn %>% filter(enzyme == "Trypsin") %>%pull(fasta_headers),
  pro_try = sct_venn %>% filter(enzyme == "ProAlanase+trypsin") %>%pull(fasta_headers))
f35_venn <- list(
  chy_try = f35_venn %>% filter(enzyme == "Chymotrypsin+trypsin") %>% pull(fasta_headers),
  glu_try = f35_venn %>% filter(enzyme == "Glu-C+trypsin") %>%pull(fasta_headers),
  try = f35_venn %>% filter(enzyme == "Trypsin") %>%pull(fasta_headers),
  pro_try = f35_venn %>% filter(enzyme == "ProAlanase+trypsin") %>%pull(fasta_headers))
grl_venn <- list(
  chy_try = grl_venn %>% filter(enzyme == "Chymotrypsin+trypsin") %>% pull(fasta_headers),
  glu_try = grl_venn %>% filter(enzyme == "Glu-C+trypsin") %>%pull(fasta_headers),
  try = grl_venn %>% filter(enzyme == "Trypsin") %>%pull(fasta_headers),
  pro_try = grl_venn %>% filter(enzyme == "ProAlanase+trypsin") %>%pull(fasta_headers))

# Plot
sct_venn_fig <- venn(sct_venn, ilables = "counts", zcolor=met.brewer("Derain", 5))
f35_venn_fig <- venn(f35_venn, ilables = "counts", zcolor=met.brewer("Derain", 5))
grl_venn_fig <- venn(grl_venn, ilables = "counts", zcolor=met.brewer("Derain", 5))
```

## Protein coverage

COL1A1 - P02453 and A0A5F5Q281.
AHSG/FETUA - P12763 and A0A5F5Q0V8.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load peptide files
pep_horse <- read_delim("<PATH_TO>/peptides.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(str_detect(proteins, "A0A5F5Q0V8")) %>%
  dplyr::select(92:113, start_position, end_position)
pep_cow <- read_delim("<PATH_TO>/peptides.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(str_detect(proteins, "P12763")) %>%
  dplyr::select(68:77, start_position, end_position)

# Fix sample names
colnames(pep_horse)[1:22] <- c("f35_chy_try1", "f35_chy_try2", "f35_chy_try3",
                               "f35_glu_try1", "f35_glu_try2", "f35_glu_try3",
                               "f35_pro_try1", "f35_pro_try2",
                               "f35_try1", "f35_try2", "f35_try3",
                               "grl_chy_try1", "grl_chy_try2", "grl_chy_try3",
                               "grl_glu_try1", "grl_glu_try2", "grl_glu_try3",
                               "grl_pro_try1", "grl_pro_try2",
                               "grl_try1", "grl_try2", "grl_try3")
colnames(pep_cow)[1:10] <- c("sct_chy_try1", "sct_chy_try2", "sct_chy_try3",
                               "sct_glu_try1", "sct_glu_try2", "sct_glu_try3",
                               "sct_pro_try1", 
                               "sct_try1", "sct_try2", "sct_try3")

# Expand numbers between start and end to get a dataframe of all covered positions
pep_horse <- pep_horse %>%
  mutate(covered = map2(start_position, end_position, seq, by=1))  
pos_list <- pep_horse %>% dplyr::select(covered) %>%
  mutate(covered = map(covered, ~as_data_frame(t(.x)))) %>% unnest()
pep_horse <- cbind(pep_horse, pos_list)

pep_cow <- pep_cow %>%
  mutate(covered = map2(start_position, end_position, seq, by=1))  
pos_list <- pep_cow %>% dplyr::select(covered) %>%
  mutate(covered = map(covered, ~as_data_frame(t(.x)))) %>% unnest()
pep_cow <- cbind(pep_cow, pos_list)

# Gather to long form
pep_horse <- pep_horse %>%
  gather("remove", "position", 26:48) %>% 
  dplyr::select(-start_position, -end_position, -covered, -remove) %>%
  filter(position > 0) %>%
  gather("sample", "intensity", 1:22) %>%
  filter(intensity > 0)

pep_cow <- pep_cow %>%
  gather("remove", "position", 14:38) %>% 
  dplyr::select(-start_position, -end_position, -covered, -remove) %>%
  filter(position > 0) %>%
  gather("sample", "intensity", 1:10) %>%
  filter(intensity > 0)

# Combine datasets
pep_combined <- full_join(pep_horse, pep_cow)

# Sum intensity per position
pep_summary <- pep_combined %>%
  group_by(position, sample) %>%
  dplyr::summarize(total_intensity = sum(intensity))

# Add metadata
pep_summary <- left_join(pep_summary, meta) %>%
  dplyr::select(position, sample, total_intensity, enzyme, bone)

# Average between replicates
pep_summary <- pep_summary %>%
  group_by(position, bone, enzyme) %>%
  dplyr::summarize(mean_intensity = mean(total_intensity))

# log2 intensity
pep_summary$intensity_log2 <- log2(pep_summary$mean_intensity)

# Combine bone and enzyme columns
pep_summary <- pep_summary %>%
  unite(sample, bone:enzyme, remove=FALSE) 

# Separate into species, cut collagen
pep_summary_cow <- pep_summary %>%
  filter(bone == "SCT9") 
  filter(position < 1217)
pep_summary_horse <- pep_summary %>%
  filter(bone %in% c("GRL12497", "F35"))
  filter(position < 1184)

# Plot
fetua_cow <- ggplot(pep_summary_cow, aes(x=position, y=sample)) +
  geom_raster(aes(fill=intensity_log2)) +
  scale_fill_gradient(low="#e9eaf2ff", high="#6e9867ff", name="Intensity (log2)") +
  xlab("") + ylab("") +
  theme_bw()

# Combine plots
col1a1 <- ggarrange(col_cow, col_horse, common.legend=TRUE, ncol=1, heights=c(3,5))
fetua <- ggarrange(fetua_cow, fetua_horse, common.legend=TRUE, ncol=1, heights=c(3,5))
coverage <- ggarrange(col1a1, fetua, labels=c("a)", "b)"), common.legend=TRUE, widths = c(2,1))

# Export figure
ggsave("<PATH_TO>/enzexp2_coverage.pdf", coverage, width=16, height=4)
```

## Peptide length

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load data
evidence_horse <- read_delim("<PATH_TO>/evidence.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "F6T323|F7BC37|P13645|P0C1U8|P00766|P00761|P35527|P35908|P00766|P04264|P07477|P35908|P13645")) %>%
  dplyr::select(proteins, length, experiment, intensity)
evidence_cow <- read_delim("<PATH_TO>/evidence.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "P00766|A0A452DI93|F1N152|P0C1U8|P00761|P07477")) %>%
  dplyr::select(proteins, length, experiment, intensity)

# Combine files
evidence <- full_join(evidence_cow, evidence_horse) %>%
  filter(intensity > 0)

# Add metadata
evidence <- left_join(evidence, meta, by=c("experiment"="sample"))

# Summarize
length_summary <- evidence %>%
  filter(!experiment == "Blank") %>%
  group_by(experiment, enzyme, bone, age) %>%
  dplyr::summarise(mean_length = weighted.mean(x=length, w=intensity, na.rm=TRUE))

# Find Box-Cox transformation of the data
MASS::boxcox(lm(mean_length~bone, data=length_summary))
MASS::boxcox(lm(mean_length~enzyme, data=length_summary))

# Checking transformation
ggqqplot(length_summary$mean_length)
hist(length_summary$mean_length) 

# Full model - start with full and decrease as needed
model1 <- lm(mean_length ~ enzyme, data = length_summary)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) 

# Null model
model0 <- lm(mean_length ~ 1, data = length_summary)

# Compare models
anova(model1, model0)

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

# Plot
length <- ggplot(length_summary, aes(x=enzyme, y=mean_length, fill=enzyme)) + 
  geom_boxplot() +
  stat_summary(fun.y=median, geom="point", shape=21, size=3.5) +
  facet_wrap(~bone) +
  expand_limits(y=0) +
  scale_fill_manual(values=met.brewer("Derain", 5), name="Enzyme") +
  xlab("") +
  ylab("Mean length weighted by intensity") +
  theme_bw() +
  theme(axis.text.x=element_blank())
```

## Hydroxyproline

```{r}
# Load and clean evidence files
evidence_cow <- read_delim("<PATH_TO>/evidence.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "P00766|A0A452DI93|F1N152|P0C1U8|P00761|P07477")) %>%
  select(sequence, experiment, intensity, modified_sequence, 
         deamidation_nq, oxidation_m, hydroxyproline, gln_pyro_glu, glu_pyro_glu) 
evidence_horse <- read_delim("<PATH_TO>/evidence.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "F6T323|F7BC37|P13645|P0C1U8|P00766|P00761|P35527|P35908|P00766|P04264|P07477|P35908|P13645")) %>%
  select(sequence, experiment, intensity, modified_sequence, 
         deamidation_nq, oxidation_m, hydroxyproline, gln_pyro_glu, glu_pyro_glu)

# Subset to hydroxyproline
hydro_cow <- evidence_cow %>%
  dplyr::select(sequence, hydroxyproline, experiment)
hydro_horses <- evidence_horse %>%
  dplyr::select(sequence, hydroxyproline, experiment)

# Combine
hydro <- full_join(hydro_cow, hydro_horses)

# Count number of P in peptide
hydro$count_p <- str_count(hydro$sequence, "P")

# Remove peptides with no P
hydro <- hydro %>% filter(count_p > 0)

# Fraction hydroxyproline
hydro$fraction_hydro <- hydro$hydroxyproline / hydro$count_p

# Mean fraction hydroxyproline by sample
hydro_summary <- hydro %>%
  group_by(experiment) %>%
  dplyr::summarize(mean_hydro = mean(fraction_hydro))

# Add metadata
hydro_summary <- left_join(hydro_summary, meta, by=c("experiment"="sample")) %>%
  filter(!experiment == "blank")

# Find Box-Cox transformation of the data
MASS::boxcox(lm(mean_hydro~bone, data=hydro_summary))
MASS::boxcox(lm(mean_hydro~enzyme, data=hydro_summary)) 

# Checking transformation
ggqqplot(hydro_summary$mean_hydro)
hist(hydro_summary$mean_hydro) 

# Full model - start with full and reduce as needed 
model1 <- lmer(mean_hydro ~ enzyme + (1|bone), data = hydro_summary)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) 

# Null model
model0 <- lmer(mean_hydro ~ 1 + (1|bone), data = hydro_summary)

# Compare models
anova(model1, model0)

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

# Plot
ggplot(hydro_summary, aes(x=enzyme, y=mean_hydro, fill=enzyme)) + 
  geom_boxplot() +
  stat_summary(fun.y=median, geom="point", shape=21, size=3.5) +
  facet_wrap(~bone) +
  expand_limits(y=0) +
  scale_fill_manual(values=met.brewer("Derain", 5), name="Enzyme") +
  xlab("") +
  ylab("") +
  theme_bw() +
  theme(axis.text.x=element_blank())
```

# Deamidation

This analysis is based on output from the deamidate.py script published by Mackie et al. 2018. Input was cleaned evidence.txt files.

```{r}
# Load results
deam <- read.csv("<PATH_TO>/enzexp2_deamidation.txt", sep="") %>%
  clean_names()

# Extract stagetip number
deam <- deam %>%
  separate(raw_file, c("remove", "stagetip"), sep="DDA_") %>%
  select(-remove)

# Add metadata
deam <- left_join(deam, meta) %>%
  filter(!bone == "blank")

# Separate into N and Q for stats
deam_n <- deam %>% filter(n_q == "N")
deam_q <- deam %>% filter(n_q == "Q")

### N

# Find Box-Cox transformation of the data
MASS::boxcox(lm(mean~bone, data=deam_n))
MASS::boxcox(lm(mean~enzyme, data=deam_n)) 

# Checking transformation
ggqqplot(log(deam_n$mean)) 
hist(log(deam_n$mean)) 

# Full model - start with full and reduce as needed
model1 <- lmer(log(mean) ~ enzyme + (1|age), data = deam_n)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) 

# Null model
model0 <- lmer(log(mean) ~ 1 + (1|age), data = deam_n)

# Compare models
anova(model1, model0) 

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

### Q

# Find Box-Cox transformation of the data
MASS::boxcox(lm(mean~bone, data=deam_q))
MASS::boxcox(lm(mean~enzyme, data=deam_q)) 

# Checking transformation
ggqqplot(sqrt(deam_q$mean)) 
hist(sqrt(deam_q$mean)) 

# Full model - start with full and reduce as needed
model1 <- lmer(sqrt(mean) ~ enzyme + (1|bone), data = deam_q)

# See if random factor is needed 
rand(model1, reduce.terms = TRUE) 

# Null model
model0 <- lmer(sqrt(mean) ~ 1 + (1|bone), data = deam_q)

# Compare models
anova(model1, model0) 

# Check model
plot(model1)
qqnorm(resid(model1))
qqline(resid(model1))

# Plot
deamidation <- ggplot(deam, aes(x=enzyme, y=mean, fill=enzyme)) + 
  geom_boxplot() +
  stat_summary(fun.y=median, geom="point", shape=21, size=3.5) +
  scale_fill_manual(values=met.brewer("Derain", 5), name="Amino acid") +
  facet_grid(n_q~bone) +
  expand_limits(y=c(0,100)) +
  xlab("") + 
  ylab("Deamidation") +
  theme_bw()  +
  theme(axis.text.x=element_blank())

# Combine plots
diagenesis <- ggarrange(length, deamidation, labels=c("a)", "b)"), common.legend=TRUE, legend="bottom", nrow=1)

# Export figure
ggsave("<PATH_TO>/enzexp2_diagenesis.pdf", diagenesis, width=8, height=4)
```

## Cleavage sites

```{r}
# Import datasets 
peptide_cow <- read_delim("<PATH_TO>/peptides.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "P00766|A0A452DI93|F1N152|P0C1U8|P00761|P07477")) %>%
  select("n_term_cleavage_window", 71:80)
peptide_horse <- read_delim("<PATH_TO>/peptides.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  clean_names() %>%
  filter(!reverse %in% c("+")) %>%
  filter(!str_detect(proteins, "F6T323|F7BC37|P13645|P0C1U8|P00766|P00761|P35527|P35908|P00766|P04264|P07477|P35908|P13645")) %>%
  select("n_term_cleavage_window", 95:116)

# Remove peptides with missing aa's
peptide_cow <- peptide_cow %>%
  filter(!grepl('_', n_term_cleavage_window))
peptide_horse <- peptide_horse %>%
  filter(!grepl('_', n_term_cleavage_window))

# Filter for  presence
peptide_cow <- peptide_cow %>%
  pivot_longer(names_to = "enzyme", values_to = "intensity", 2:11) %>%
  filter(intensity >0) %>%
  separate(enzyme, c("remove", "sample"), "sity_") %>%
  select(-remove)
peptide_horse <- peptide_horse %>%
  pivot_longer(names_to = "enzyme", values_to = "intensity", 2:23) %>%
  filter(intensity >0) %>%
  separate(enzyme, c("remove", "sample"), "sity_") %>%
  select(-remove)

# Cut first and last 10 aa's
peptide_cow <- peptide_cow %>%
  mutate(n_term_cleavage_window = str_sub(n_term_cleavage_window, 11, -1)) %>%
  mutate(n_term_cleavage_window = str_sub(n_term_cleavage_window, end=-11))
peptide_horse <- peptide_horse %>%
  mutate(n_term_cleavage_window = str_sub(n_term_cleavage_window, 11, -1)) %>%
  mutate(n_term_cleavage_window = str_sub(n_term_cleavage_window, end=-11))

# Add metadata
peptide_cow <- left_join(peptide_cow, meta)
peptide_horse <- left_join(peptide_horse, meta)

# Summarize
peptide_cow_summary <- peptide_cow %>%
  select(n_term_cleavage_window, enzyme, bone) %>%
  distinct()
peptide_horse_summary <- peptide_horse %>%
  select(n_term_cleavage_window, enzyme, bone) %>%
  distinct()

# Make into list - there's gotta be a more smooth way to do this but I'm still tired
peptide_cow_trypsin <- peptide_cow_summary %>%
  filter(enzyme == "Trypsin")
peptide_cow_chy <- peptide_cow_summary %>%
  filter(enzyme == "Chymotrypsin+trypsin")
peptide_cow_glu <- peptide_cow_summary %>%
  filter(enzyme == "Glu-C+trypsin")
peptide_cow_pro <- peptide_cow_summary %>%
  filter(enzyme == "ProAlanase+trypsin")
peptide_cow_list <- list(peptide_cow_trypsin$n_term_cleavage_window,
                       peptide_cow_chy$n_term_cleavage_window,
                       peptide_cow_glu$n_term_cleavage_window,
                       peptide_cow_pro$n_term_cleavage_window)
names(peptide_cow_list) <- c("Trypsin", "Chymotrypsin+trypsin", "Glu-C+trypsin", "ProAlanase+trypsin") 

peptide_horse_trypsin <- peptide_horse_summary %>%
  filter(enzyme == "Trypsin")
peptide_horse_chy <- peptide_horse_summary %>%
  filter(enzyme == "Chymotrypsin+trypsin")
peptide_horse_glu <- peptide_horse_summary %>%
  filter(enzyme == "Glu-C+trypsin")
peptide_horse_pro <- peptide_horse_summary %>%
  filter(enzyme == "ProAlanase+trypsin")
peptide_horse_list <- list(peptide_horse_trypsin$n_term_cleavage_window,
                       peptide_horse_chy$n_term_cleavage_window,
                       peptide_horse_glu$n_term_cleavage_window,
                       peptide_horse_pro$n_term_cleavage_window)
names(peptide_horse_list) <- c("Trypsin", "Chymotrypsin+trypsin", "Glu-C+trypsin", "ProAlanase+trypsin") 

# Plot
seq_cow <- ggseqlogo(peptide_cow_list, ncol=2)
seq_horse <- ggseqlogo(peptide_horse_list, ncol=2)

# Combine plots
seq_logos <- ggarrange(seq_cow, seq_horse, nrow=1, common.legend=TRUE, legend="bottom",
          labels=c("a)", "b)"))
```
